Testing for measurement invariance with respect to an ordinal variable

### 2013-04-17 ordinal/response.Rnw

\documentclass[a4paper]{article}
\usepackage{a4wide}

\begin{document}

\setkeys{Gin}{width=\textwidth}

\section*{Reviewer 1}

\begin{enumerate}
\item \emph{A comparison of the current approach with scenarios in which the
factor analyst uses Modification Indices (MIs) to free parameters. In practice
if one has, say, 5 age groups and a given parameter is found to lead to
significant increases in model fit when restricted across groups, it is quite
common to use MIs to detect the source(s) of the misfit and then to end up with,
say one parameter for age groups 1-3  and another parameter for age groups 4 \& 5
in a so-called partially invariant model (see Byrne et al., 1989). It would be
interesting to see how that approach fares in comparison to the approaches
described in the manuscript. I am also interested in reading the authors' ideas
about the use MI in ``ad hoc" model fitting.}

MIs are quite related to the proposed tests: the MI described by, e.g., Sorbom
1989 is equivalent to a Lagrange multiplier test.  The maximum of Lagrange
multiplier tests (e.g., computed across all potential thresholds of V), on the
other hand, is within the family of statistics described by Merkle \& Zeileis
(2013).  However, as the reviewer states, MIs are typically applied only after a
significant likelihood ratio test has been obtained.  Our proposed statistics
can be used across multiple parameters, potentially in place of the likelihood
ratio test.  If the test is significant, one could use MIs or parameter
fluctuation plots such as those in manuscript Figures 3 \& 4 to determine the
model misfit.  We have added some brief detail on this relationship to the
"Tests for Continuous V" section.

\item \emph{Given the widespread literature on ordinal factor analyses, it would
be worthwhile to link the current modelling approach to ordinal factor analysis.
Another major reason is that ordinal factor analysis naturally extends towards
more standard IRT models. Therefore, by bridging the gap from the current
modelling approach to ordinal factor analysis, the authors may also take steps
to integrate it with the literature in measurement invariance in the IRT
framework.}

We have added detail in the General Discussion on the extension of the proposed
tests to other factor analysis models and to IRT.  In particular, our proposed
tests can potentially extend to any psychometric model that assumes independent
observations and that uses ML estimation.  Additionally, we cite some work on
the application of similar tests to IRT.

\item \emph{Study of power under other violations of MI related to factor
loadings or measurement intercepts.}

We now acknowledge that the obtained simulation results may not generalize to
other measurement invariance violations or to other factor analysis models.  We
feel that this suggestion would require multiple simulations that would result
in a much longer paper, so we have not added new simulations to the manuscript.

\item \emph{Consider issues related to model comparison by means of AICs or BICs
and the like. This is particularly relevant in light of the fact that monotonic
violations of MI should be detectable in a more parsimonious manner than how it
is often done in practice with likelihood ratio tests with many DFs.}

If we understand this comment correctly, the reviewer is suggesting that we
compare the proposed statistics to the use of AIC or BIC (in addition to the
LRT).  Under this framework, one would, e.g., reject the hypothesis of
measurement invariance if the AIC for the ``full" model were less than the AIC
for the restricted model.  We have added the AIC strategy to our simulation,
along with some other adjusted versions of the LRT (see Response 2.2).  The AIC
turns out to have particularly low power.
\end{enumerate}

\section*{Reviewer 2}

\begin{enumerate} \item \emph{On page 5 is the scaled sum of scores which is
normalized by the square root of n and the covariance matrix of the scores I.
This is fine theoretically, but what choice do you make in practice? In fact I
think that here the problem is model misspecification, other than the biasing
factor V. If the model used in ML is incorrect with respect to the number of
factors for instance, then this will result in a inconsistent estimate of I. Why
not use the sandwich by the late Halbert White (JASA, 1982 for instance)? This
is easy to calculate and will improve the limiting distribution of the test
statistics.  The choice of I is also missing in the simulations. Please be more
specific what you did there.}

We have added an extended discussion of choices for the covariance matrix of
scores, with references.  We have also clarified that, in the simulations, we
used the information matrix.  Because our simulation models were correctly
specified (except for the measurement invariance violation), we do not expect
the choice of I to play a large role there.

\item \emph{The simulations show great improvement over the standard LR test on
measurement invariance. And on page 10 the authors mention the results by
Bentler and Bonett (1980) on LR tests picking up deviations too quickly. I think
the alternative tests, adjustments of LR tests, like the one proposed in Yuan
and Bentler (JASA, 1997). There the LR is downweighted at rate $O_p(1/n)$, which
brings the results of the LR closer to the results of the authors' maxLM test. I
would like to see a comparison with this or maybe some other adjusted LR test.}

<<fig1, echo=FALSE, results=hide, fig=TRUE, height=10, width=10>>=
load("sim1.rda")
library("lattice")

lev0 <- c(   "LM_o","LRT","LRT_sb","AIC")
lev  <- c("maxLM_o","LRT","LRT_sb","AIC")
cl <- c(1, 4, 2, 3)
pc <- c(1, 2, 3, 4)
lt <- c(1, 1, 2, 2)

sim1.tmp <- sim1[sim1$test %in% lev0,]
sim1.tmp$test <- factor(sim1.tmp$test, levels = lev0, labels = lev)
trellis.par.set(theme = canonical.theme(color = FALSE))
print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim1.tmp,
  type = "b", xlab="Violation Magnitude", ylab="Power",
  key=list(text = list(lab = lev), lines = list(col = cl, lty = lt), points = list(col = cl, pch = pc)),
  col = cl, pch = pc, lty = lt))
@

<<fig2, echo=FALSE, results=hide, fig=TRUE, height=10, width=10>>=
load("sim2.rda")
sim2.tmp <- sim2[sim2$test %in% lev0,]
sim2.tmp$test <- factor(sim2.tmp$test, levels = lev0, labels = lev)
trellis.par.set(theme = canonical.theme(color = FALSE))
print(xyplot(power ~ diff | nlevels + nobs, group = ~ test, data = sim2.tmp,
  type = "b", xlab="Violation Magnitude", ylab="Power",
  key=list(text = list(lab = lev), lines = list(col = cl, lty = lt), points = list(col = cl, pch = pc)),
  col = cl, pch = pc, lty = lt))
@

We implemented three adjusted LRTs in our simulations: (i) the Yuan-Bentler 1997
adjustment suggested by the reviewer, (ii) the use of AIC, and (iii) the
Satorra-Bentler adjustment with correction for difference tests.  While these
adjustments downweight the LRT associated with individual models (e.g.,
comparing an individual model to a saturated model), the impact of the
adjustments on difference tests is less clear.  Further, the Yuan-Bentler
adjustment is seemingly designed for situations where the model is estimated via
WLS, and there are no available corrections for difference tests.  We suspect
that these issues are problematic, as our simulation results indicated that
these statistics are worse than the usual LRT in nearly all conditions.
We have included the resulting graphs for Simulation~1 and Simulation~2 above
(but not in the manuscript).
\begin{itemize}
  \item Both graphs are organized in the same manner as the graphs in the paper,
    except that they contain different statistics: the ordinal maxLM test
    ($\mathrm{LM}_o$), the Likelihood ratio test (LRT), the Satorra-Bentler LRT ($\mathrm{LRT}_\mathrm{sb}$),
    and the use of AIC (AIC). The former two are presented in the manuscripts while the latter
    two are only shown here.
  \item It can be seen that $\mathrm{LRT}_\mathrm{sb}$ is virtually identical to
    the plain LRT in situations with small number of groups $m$ and/or large
    sample sizes $n$ (in particular throughout Simulation~2) while for the large $m$/small
    $n$ scenarios , the problems of the LRT are even amplified.
  \item Finally, the AIC has relatively low power throughout Simulation~1. This is
    because using AIC corresponds to using the LRT with an increased critical value
    here (because the models being compared are nested with relatively large differences in degrees
    of freedom).
\end{itemize}
Hence, we feel that the additional discussion of these modifications do not add
substantial new insights and hence have omitted them from the paper (and only
mention them very briefly). But we are willing to reconsider if the AE or
reviewers find the results worthwhile.
\end{enumerate}

\section*{Reviewer 3}

\begin{enumerate} \item \emph{The section entitled `Theoretical Detail' (page 4)
starts with the background for the score-based test statistics. However, from
lines 18 onwards, there is a section on measurement invariance (starting with
`In studying measurement invariance, we are essentially testing...' up to line
19 on page 5, just before the subsection `Tests for Continuous V'). I would
suggest to put the section on measurement invariance testing in a separate
(sub)section, and present this first, before providing the background for the
test statistics.}

We have moved around these sections, so that measurement invariance comes before
the theoretical discussion.

\item \emph{For readers unfamiliar with case-based scores, it may be useful to
have a table with artificial values for some $B(\theta)_{ij}$ values; say for 20
subjects, and 4 parameters; in an additional column, there could be ordered
continuous values for a V variable; in the next column, the V values could be
binned to get an ordinal variable; referring to this table, it would be trivial
to illustrate how the values of eg DM, CvM, maxLM, etc are computed...}

We agree that this would be useful for some readers, but we think that it would
take up more space than it is worth.  For example, the calculation of ordinal
test statistics would require both calculations within each bin and unique
scaling (by $(i/n) \cdot (1 - i/n)$) for each bin.  While these calculations are not
difficult, they also consume space to write them out (especially if we included
the non-ordinal statistics).  Additionally, besides the calculation of test
statistics, it might be equally useful to show how the scores $s(\theta; x_i)$ can
be used to calculate $B(\theta)$.  This implies that we would need both a table of
$s(\theta; x_i)$ and a table of cumulative scores $B(\theta)$.  We feel that these
ideas are represented more concisely and intuitively via their equations.

\item \emph{related to 2), I would suggest adding (perhaps in an appendix) some
R code illustrating how (for example, for the values in the aforementioned
table) the p-values can be computed for the various test statistics}

We have added some practical discussion of p-value computation to the ``Choice of
Test" subsection of the General Discussion.  We have also posted all the R code
for replicating our results (p-values and otherwise) on our R-Forge page (listed
in the ``Computational Details" section).  We do not include code within the
manuscript because the p-values associated with some of the statistics are
computed via simulation (i.e., there is not a closed-form solution for critical
values associated with the test statistic), and much of the underlying R code is
already contained within R's strucchange package.
\end{enumerate}

All minor points have been addressed, with thanks to the reviewer.

\end{document}

### 2013-04-22 ordinal/ordinal.Rnw
Final to ordinal ms prior to resubmission o Added YB97 statistic to response o Used apa6 class instead of apa class for the ms,   allowing for easy masking of authors
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@200 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Ordinal ms, unmasked the author names
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@201 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Final read of ordinal ms, minor wording changes.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@202 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[man]{apa}
\usepackage{graphicx,epsfig,amsmath,alltt,setspace,bm}
%% This keeps the tildes out when you use Rnw2pdf:
\usepackage[english]{babel}
%% need no \usepackage{Sweave}
\SweaveOpts{engine = R, eps = FALSE, echo = FALSE, results = hide}

@@

\acknowledgements{This work was supported by National Science
  Foundation grant SES-1061334. 
  Correspondence to Edgar C.\ Merkle, Department of  
  Psychological Sciences, University of Missouri, Columbia,
  MO 65211.
  Email: {\texttt{merklee@missouri.edu}}.}
\shorttitle{Ordinal measurement invariance}
\rightheader{Ordinal measurement invariance}

@@

The study of measurement invariance and differential item functioning
(DIF) 
has received considerable attention in the psychometric literature
(see, e.g., \citeNP{Mil11} for a thorough review).  A set of
psychometric scales $X$ are defined to be measurement invariant
with respect to an auxiliary variable $V$ if \cite{Mel89}
\begin{equation}
    \label{eq:midef}
      f(x_i | t_i, v_i, \dots) = f(x_i | t_i, \dots),
\end{equation}
where $T$ is the latent variable that the scales measure, $f$ is the
model's distributional form, the $i$ subscript refers to
individual cases, capital letters signify random variables, and lowercase
letters signify realizations of the variables.  If the above equation
does not hold, then  
a measurement invariance violation is said to exist.  We focus here on
situations where $f()$ is the probability density function of $X$,
and the measurement invariance
violation occurs because the model parameters are
unequal across individuals (and related to $V$).

@@

Along with categorical variables such as ethnicity, researchers are
often interested in studying 
measurement invariance with respect to ordinal $V$.
Such variables can arise from multiple choice surveys, where
continuous variables such as age or income are binned into a small number
of categories.  Alternatively, the variables may arise from gross,
qualitative assessments of a particular measure of interest, where
individuals may be categorized as having a
``low,'' ``medium,'' or ``high'' level of the variable of interest.
While these variables are relatively easy to find in the literature,
there exist very few psychometric methods that specifically
account for the fact that $V$ is ordinal.  More often, $V$ is 
treated as categorical so that the traditional tests can be applied.  Alternatively, if there are many levels of $V$, $V$ may also be treated as continuous. 
It is the intent of this paper to propose
two test statistics that explicitly treat $V$ as ordinal and to show
that the statistics 
possess good properties for use in practice.

@@

Tests of  measurement invariance are essentially tests of the 
assumption that all  
individuals arise from the same 
parameter vector ${\bm \theta}$.
Thus, a hypothesis of measurement invariance can be written as
\begin{equation}
    \label{eq:h0}
    H_0:~ {\bm \theta}_i = {\bm \theta}_0,\quad (i=1,\ldots,n),
\end{equation}
where ${\bm \theta}_i$ reflects the parameter vector for individual
$i$ (and modifications for subsets of ${\bm \theta}$ are immediate).  
The most general alternative hypothesis related to $V$ may then be
written as
\begin{equation}
  \label{eq:h2*}
  H_1^*:~ {\bm \theta}_i = {\bm \theta}_{v_i},
\end{equation}
stating that the parameter vector differs for every unique realization
of $V$.  This alternative is commonly employed when $V$ is
categorical.  In these situations, the likelihood ratio test (LRT) compares
a model where 
parameters are restricted across groups (i.e., across values of $V$)
to a model where parameters are free across groups; the exact
parameter values within each group are completely unrestricted.
However, in situations where $V$ is categorical or
continuous,~\eqref{eq:h2*} includes non-monotonic violations of
measurement invariance.   This allows for instances where, e.g., the
parameter values initially increase with $V$ and then decrease, 
or where just one or two ``middle'' levels of $V$ differ from the rest.
Researchers typically do not expect such a result when testing
measurement invariance w.r.t.\ categorical or continuous $V$, and 
researchers often cannot interpret such violations.  Monotonic
parameter changes w.r.t.\ $V$ are of much more interest in these
situations, with the simplest type of change given by the alternative
hypothesis
\begin{equation}
  \label{eq:h1*}
  H_2^*:~ {\bm \theta}_i = \left\{ \begin{array}{ll}
    {\bm \theta}^{(A)} & \mbox{if } v_i \le \nu, \\
    {\bm \theta}^{(B)} & \mbox{if } v_i >   \nu,
  \end{array} \right.
\end{equation}
where $\nu$ is a threshold dividing individuals into two groups based on
$V$.  This alternative is implicitly employed in ``median split''
analyses, 
where $\nu$ is given as the sample median of $V$.  The threshold $\nu$
is usually unknown, however, so it is generally of interest to
test~\eqref{eq:h1*} across all possible values of $\nu$.  The tests
proposed below generally allow for this.

As stated previously, we specifically focus here on situations where $V$ is
ordinal and where the measurement invariance violation is related to the ordinal
variable (e.g., the violation is of the type from~\eqref{eq:h1*} or
the violation grows/shrinks with $V$).  Researchers typically test for
measurement invariance w.r.t.\ ordinal $V$ by employing the alternative
from~\eqref{eq:h2*}, which implicitly treats $V$ as categorical.
Thus, test statistics that explicitly treat $V$ as ordinal should have
higher power to detect measurement invariance violations that are
monotonic with $V$.

@@

To formalize the ideas discussed in the previous paragraph, we assume
that the observations are ordered 
w.r.t.\ $V$, with $x_{(i)}$ reflecting
the data for the individual who has the 
$i^{\text{th}}$-smallest value of $V$.  We then define the $k$-dimensional
{\em cumulative score process} as 
\begin{equation} \label{eq:cumscore}
  {\bm B}(t; \hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
    \sum_{i = 1}^{\lfloor n t \rfloor} {\bm s}(\hat {\bm \theta}; x_{(i)})
  \qquad (0 \le t \le 1)
\end{equation}
where $\lfloor nt \rfloor$ is the integer part of $nt$ and
$\hat I$ is some consistent estimate of the covariance matrix of the
scores.  Natural choices for $\hat I$ include the information matrix
(which we use in our applications and simulations below) or
alternatively some kind of outer product of the scores or sandwich
estimator to guard the inference against potential misspecification of
the model (see \citeNP{hub67} for the theoretical foundation and
\citeNP{zei06b} for a computational framework).  
Equation~\eqref{eq:cumscore} simultaneously accounts for the
ordering of individuals w.r.t.\ $V$ and decorrelates the scores
associated with each of the $k$ model parameters (which allows us to
potentially make inferences separately for each individual model
parameter).  Using ideas similar to those that were outlined in the previous
paragraph, the cumulative score process associated with each
model parameter should randomly fluctuate around zero under
measurement invariance.  Further, there exists a functional central
limit theorem that allows us to make formal inference with this
cumulative score process.  Assuming that individuals are independent
and the usual ML regularity conditions hold, it is possible to show
that \cite{HjoKon02}
\begin{equation} \label{eq:fclt}
  {\bm B}(\cdot; \hat {\bm \theta}) ~\overset{d}{\rightarrow}~ {\bm B}^{0}(\cdot),
\end{equation}
where $\overset{d}{\rightarrow}$ denotes convergence in distribution
and ${\bm B}^{0}(\cdot)$ is a $k$-dimensional Brownian bridge.
Thus, we can construct tests of measurement invariance by comparing
the behavior of the cumulative score process to that of a Brownian bridge.
This is accomplished by comparing a scalar statistic associated with
the cumulative score process to the analogous statistic of a Brownian
bridge.

@@

If $V$ is only nominal/categorical,
there is not even a partial ordering, i.e., measurement invariance tests should neither
exploit the ordering of $V$'s levels nor of the observations within the level.
In this situation, it is possible to obtain a test statistic by first summing scores within
each of the $m$ levels of the auxiliary variable, then ``summing the sums'' to obtain a test
statistic \cite{HjoKon02}.  This test statistic can be formally written as
\begin{equation}
    \label{eq:lmuo}
     \mathit{LM}_\mathit{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
       \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}(\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2,
\end{equation}
where, again, tests of subsets of model parameters can be obtained by
taking the inner sum over only the $k^* < k$ parameters of interest.
This test statistic discards the ordinal nature of the auxiliary
variable, essentially employing the alternative hypothesis
from~\eqref{eq:h2*}.  A similar issue is observed in testing for
measurement 
invariance via multiple groups models and likelihood ratio tests (or,
equivalently, via Wald tests or Lagrange multiplier tests): we
can allow $\bm{\theta}$ to be unique at each level of the ordinal
variable, but the ordinality of the auxiliary variable is lost.
In contrast, the statistics proposed above explicitly account for the fact
that $V$ is ordinal.

As demonstrated in the simulations below, the proposed ordinal test
statistics are sensitive to the measurement invariance
violations that an analyst would typically expect from an ordinal $V$.
In particular, due to computing cumulative sums in ${\bm B}(\hat {\bm \theta})$,
violations that occur as we move along the levels of $V$ can be captured
well. This includes abrupt shifts in the parameters $\bm \theta$ at a
certain level of $V$ as well as smooth increases/decreases in the parameters.
Taking a maximum over the $k$~parameters as in $\mathit{WDM}_o$ will be
more sensitive to changes that occur only in one out of many
parameters, 
while $\max \mathit{LM}_o$ will be more sensitive to changes occurring
in several (or even all of the) parameters simultaneously.
Moreover, the test statistics are rather insensitive to anomalies in a small
number of categories of $V$ that are
unrelated to the ordering of $V$.  This is especially relevant to situations
in which the analyst has a large sample size, so that the usual
likelihood ratio test is notoriously sensitive to minor parameter
instabilities (e.g., \citeNP{Ben80}).

@@

The two ``traditional'' statistics that we consider generally treat
the ordinal auxiliary variable as categorical.  These include the
likelihood ratio test of measurement 
invariance in the six unique variance parameters and the
unordered LM test from~\eqref{eq:lmuo}.  At the request of reviewers,
we also considered the
Satorra-Bentler \citeyear{SatBen01} scaled likelihood ratio test
statistic with correction for difference testing, the Yuan-Bentler
\citeyear{YuaBen97} scaled test statistic, and the use of AIC
\cite{Aka74} for detecting measurement 
invariance.  We do not report the latter results, because these test
statistics performed worse than the usual likelihood ratio test (for
both Simulation 1 and Simulation 2).

@@

Simulation results comparing the ordinal tests to the unordered LM
test and the LRT are presented in Figure~\ref{fig:sim1res}.  Rows of
the figure 
correspond to $n$, columns of the figure correspond to $m$, the x-axis
of each panel corresponds to $d$, and the y-axis of each panel
corresponds to power.  It is seen that one of the proposed test
statistics, the $\max \mathit{LM}_o$ statistic
from~\eqref{eq:maxlmo},
generally has the largest power to detect the ordinal measurement
invariance violations.  The other three tests are considerably closer
in power, with the second proposed ordinal statistic
(the double-max test from~\eqref{eq:wdm}) exhibiting the lowest power
at large violation magnitudes.  This is because the double-max
test discards information about multiple parameters changing together
at specific levels of the ordinal variable
(see \citeNP{MerZei13}, for related discussion), while the
three other tests under consideration make use of this information.
Finally, it is seen that, in the small $n$ and large $m$ conditions,
the likelihood ratio test exhibits large Type-I error rates (i.e.,
power greater than 0.05 at $d=0$).  This is because the likelihood
ratio test requires estimation of a multiple-groups model, which is very
unstable with large numbers of groups and small sample sizes as only
$n/m$ observations are available in each subsample.  The proposed
statistics are all of the LM-type and just require estimation of the
single-group model, 
leading to a clear advantage in these conditions.

@@

In this simulation, we demonstrate that the proposed statistics are 
relatively insensitive to minor parameter violations that are
unrelated to the ordering of the auxiliary variable.  As noted
earlier, this feature is especially applicable to situations where
one's sample size is very large.  Analysts often resort to informal
fit measures in this case, because the traditional LRT is nearly
guaranteed to result in significance.  This simulation is intended to
show that the proposed ordinal tests remain viable for large $n$.

@@

\citeA{FroFan11} used confirmatory factor 
models to study the invariance of three youth gratitude scales across
students aged 10 to 19 years.  Due to sample size constraints, the age
variable included six categories: 10--11 years, 12--13 years, 14 years,
15 years, 16 years, and 17--19 years.  Thus, age is an ordinal variable
with which the proposed tests can be applied.

>>>>

\documentclass[man,apacite]{apa6}
\usepackage{graphicx,epsfig,amsmath,alltt,setspace,bm}
%% This keeps the tildes out when you use Rnw2pdf:
\usepackage[english]{babel}
%% need no \usepackage{Sweave}
\SweaveOpts{engine = R, eps = FALSE, echo = FALSE, results = hide}

@@

\authornote{This work was supported by National Science
  Foundation grant SES-1061334. 
  Correspondence to Edgar C.\ Merkle, Department of  
  Psychological Sciences, University of Missouri, Columbia,
  MO 65211.
  Email: {\texttt{merklee@missouri.edu}}.}
\shorttitle{Ordinal measurement invariance}
\rightheader{Ordinal measurement invariance}

@@

The study of measurement invariance and differential item functioning
(DIF) 
has received considerable attention in the psychometric literature
(see, e.g., \citeNP{Mil11} for a thorough review).  A set of
psychometric scales $X$ is defined to be measurement invariant
with respect to an auxiliary variable $V$ if \cite{Mel89}
\begin{equation}
    \label{eq:midef}
      f(x_i | t_i, v_i, \dots) = f(x_i | t_i, \dots),
\end{equation}
where $T$ is the latent variable that the scales measure, $f$ is the
model's distributional form, the $i$ subscript refers to
individual cases, capital letters signify random variables, and lowercase
letters signify realizations of the variables.  If the above equation
does not hold, then  
a measurement invariance violation is said to exist.  We focus here on
situations where $f()$ is the probability density function of $X$,
and the measurement invariance
violation occurs because the model parameters are
unequal across individuals (and related to $V$).

@@

Along with categorical variables such as ethnicity, researchers are
often interested in studying 
measurement invariance with respect to ordinal $V$.
Such variables can arise from multiple choice surveys, where
continuous variables such as age or income are binned into a small number
of categories.  Alternatively, the variables may arise from gross,
qualitative assessments of a particular measure of interest, where
individuals may be categorized as having a
``low,'' ``medium,'' or ``high'' level of the variable of interest.
While these variables are relatively easy to find in the literature,
there exist very few psychometric methods that specifically
account for the fact that $V$ is ordinal.  More often, $V$ is 
treated as categorical so that the traditional tests can be applied.
Additionally, if there are many levels, then $V$ may also be treated
as continuous.  
The goals of this paper are to propose 
two test statistics that explicitly treat $V$ as ordinal and to show
that the statistics 
possess good properties for use in practice.

@@

Tests of  measurement invariance are essentially tests of the 
assumption that all  
individuals arise from the same 
parameter vector ${\bm \theta}$.
Thus, a hypothesis of measurement invariance can be written as
\begin{equation}
    \label{eq:h0}
    H_0:~ {\bm \theta}_i = {\bm \theta}_0,\quad (i=1,\ldots,n),
\end{equation}
where ${\bm \theta}_i$ reflects the parameter vector for individual
$i$ (and modifications for subsets of ${\bm \theta}$ are immediate).  
The most general alternative hypothesis related to $V$ may then be
written as
\begin{equation}
  \label{eq:h2*}
  H_1^*:~ {\bm \theta}_i = {\bm \theta}_{v_i},
\end{equation}
stating that the parameter vector differs for every unique realization
of $V$.  This alternative is commonly employed when $V$ is
categorical.  In these situations, the likelihood ratio test (LRT) compares
a model where 
parameters are restricted across groups (i.e., across values of $V$)
to a model where parameters are free across groups; the exact
parameter values within each group are completely unrestricted.
However, in situations where $V$ is ordinal or
continuous,~\eqref{eq:h2*} includes non-monotonic violations of
measurement invariance.   This allows for instances where, e.g., the
parameter values initially increase with $V$ and then decrease, 
or where just one or two ``middle'' levels of $V$ differ from the rest.
Researchers typically do not expect such a result when testing
measurement invariance w.r.t.\ ordinal or continuous $V$, and 
researchers often cannot interpret such violations.  Monotonic
parameter changes w.r.t.\ $V$ are of much more interest in these
situations, with the simplest type of change given by the alternative
hypothesis
\begin{equation}
  \label{eq:h1*}
  H_2^*:~ {\bm \theta}_i = \left\{ \begin{array}{ll}
    {\bm \theta}^{(A)} & \mbox{if } v_i \le \nu, \\
    {\bm \theta}^{(B)} & \mbox{if } v_i >   \nu,
  \end{array} \right.
\end{equation}
where $\nu$ is a threshold dividing individuals into two groups based on
$V$.  This alternative is implicitly employed in ``median split''
analyses, 
where $\nu$ is given as the sample median of $V$.  The threshold $\nu$
is usually unknown, however, so it is generally of interest to
test~\eqref{eq:h1*} across all possible values of $\nu$.  The tests
proposed below generally allow for this.

As stated previously, we specifically focus on situations where $V$ is
ordinal and where the measurement invariance violation is related to the ordinal
variable (e.g., the violation is of the type from~\eqref{eq:h1*} or
the violation grows/shrinks with $V$).  Researchers typically test for
measurement invariance w.r.t.\ ordinal $V$ by employing the alternative
from~\eqref{eq:h2*}, which implicitly treats $V$ as categorical.
Thus, test statistics that explicitly treat $V$ as ordinal should have
higher power to detect measurement invariance violations that are
monotonic with $V$.

@@

To formalize the ideas discussed in the previous paragraph, we assume
that the observations are ordered 
w.r.t.\ $V$, with $x_{(i)}$ reflecting
the data for the individual who has the 
$i^{\text{th}}$-smallest value of $V$.  We then define the $k$-dimensional
{\em cumulative score process} as 
\begin{equation} \label{eq:cumscore}
  {\bm B}(t; \hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
    \sum_{i = 1}^{\lfloor n \cdot t \rfloor} {\bm s}(\hat {\bm \theta}; x_{(i)})
  \qquad (0 \le t \le 1)
\end{equation}
where $\lfloor nt \rfloor$ is the integer part of $nt$ and
$\hat I$ is some consistent estimate of the covariance matrix of the
scores.  Natural choices for $\hat I$ include the information matrix
(which we use in our applications and simulations below) or
alternatively some kind of outer product of the scores or sandwich
estimator to guard the inference against potential misspecification of
the model (see \citeNP{hub67} for the theoretical foundation and
\citeNP{zei06b} for a computational framework).  
Equation~\eqref{eq:cumscore} simultaneously accounts for the
ordering of individuals w.r.t.\ $V$ and decorrelates the scores
associated with each of the $k$ model parameters (which allows us to
potentially make inferences separately for each individual model
parameter).  Using ideas similar to those that were outlined in the previous
paragraph, the cumulative score process associated with each
model parameter should randomly fluctuate around zero under
measurement invariance.  Further, there exists a functional central
limit theorem that allows us to make formal inference with this
cumulative score process.  Assuming that individuals are independent
and the usual ML regularity conditions hold, it is possible to show
that \cite{HjoKon02}
\begin{equation} \label{eq:fclt}
  {\bm B}(\cdot; \hat {\bm \theta}) ~\overset{d}{\rightarrow}~ {\bm B}^{0}(\cdot),
\end{equation}
where $\overset{d}{\rightarrow}$ denotes convergence in distribution
and ${\bm B}^{0}(\cdot)$ is a $k$-dimensional Brownian bridge.
Thus, we can construct tests of measurement invariance by comparing
the behavior of the cumulative score process to that of a Brownian bridge.
This is accomplished by comparing a scalar statistic associated with
the cumulative score process to the analogous statistic of a Brownian
bridge.

@@

If $V$ is only nominal/categorical,
there is not even a partial ordering, i.e., measurement invariance tests should neither
exploit the ordering of $V$'s levels nor of the observations within the level.
In this situation, it is possible to obtain a test statistic by first summing scores within
each of the $m$ levels of the auxiliary variable, then ``summing the sums'' to obtain a test
statistic \cite{HjoKon02}.  This test statistic can be formally written as
\begin{equation}
    \label{eq:lmuo}
     \mathit{LM}_\mathit{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
       \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}(\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2,
\end{equation}
where the first subscript on the two terms in parentheses are $i_\ell$
and $i_{\ell-1}$, respectively.  Again, tests of subsets of model
parameters can be obtained by 
taking the inner sum over only the $k^* < k$ parameters of interest.
This test statistic discards the ordinal nature of the auxiliary
variable, essentially employing the alternative hypothesis
from~\eqref{eq:h2*}.  A similar issue is observed in testing for
measurement 
invariance via multiple groups models and likelihood ratio tests (or,
equivalently, via Wald tests or Lagrange multiplier tests): we
can allow $\bm{\theta}$ to be unique at each level of the ordinal
variable, but the ordinality of the auxiliary variable is lost.
In contrast, the statistics proposed above explicitly account for the fact
that $V$ is ordinal.

As demonstrated in the simulations below, the proposed ordinal test
statistics are sensitive to the measurement invariance
violations that an analyst would typically expect from an ordinal $V$.
In particular, due to computing cumulative sums in ${\bm B}(\hat {\bm \theta})$,
violations that occur as we move along the levels of $V$ can be captured
well. This includes abrupt shifts in the parameters $\bm \theta$ at a
certain level of $V$ as well as smooth increases/decreases in the parameters.
Taking a maximum over the $k$~parameters as in $\mathit{WDM}_o$ will be
more sensitive to changes that occur only in one out of many
parameters, 
while $\max \mathit{LM}_o$ will be more sensitive to changes occurring
in several (or even all of the) parameters simultaneously.
Moreover, the test statistics are rather insensitive to anomalies in a small
number of categories of $V$ that are
unrelated to the ordering of $V$.  This is especially relevant to situations
in which the analyst has a large sample size, so that the usual
likelihood ratio test is overly sensitive to minor parameter
instabilities (e.g., \citeNP{Ben80}).

@@

The two ``traditional'' statistics that we consider generally treat
the ordinal auxiliary variable as categorical.  These include the
likelihood ratio test of measurement 
invariance in the six unique variance parameters and the
unordered LM test from~\eqref{eq:lmuo}.  At the request of reviewers,
we also considered the
Satorra-Bentler \citeyear{SatBen01} scaled likelihood ratio test
statistic with correction for difference testing, the Yuan-Bentler
\citeyear{YuaBen97} scaled test statistic, and the use of AIC
\cite{Aka74} for detecting measurement 
invariance.  We do not report the latter results, because these test
statistics performed worse than the usual likelihood ratio test (both
here and in Simulation 2).

@@

Simulation results comparing the ordinal tests to the unordered LM
test and the LRT are presented in Figure~\ref{fig:sim1res}.  Rows of
the figure 
correspond to $n$, columns of the figure correspond to $m$, the x-axis
of each panel corresponds to $d$, and the y-axis of each panel
corresponds to power.  It is seen that one of the proposed test
statistics, the $\max \mathit{LM}_o$ statistic
from~\eqref{eq:maxlmo},
generally has the largest power to detect the ordinal measurement
invariance violations.  The other three tests are considerably closer
in power, with the second proposed ordinal statistic
(the double-max test from~\eqref{eq:wdm}) exhibiting the lowest power
at large violation magnitudes.  This is because the double-max
test discards information about multiple parameters changing together
at specific levels of the ordinal variable
(see \citeNP{MerZei13}, for related discussion), while the
three other tests under consideration make use of this information.
Finally, it is seen that, in the small $n$ and large $m$ conditions,
the likelihood ratio test exhibits large Type-I error rates (i.e.,
power greater than 0.05 at $d=0$).  This is because the likelihood
ratio test requires estimation of a multiple-groups model, which is very
unstable with large numbers of groups and small sample sizes (as only
$n/m$ observations are available in each subsample).  The statistics
proposed here
are all of the LM-type and just require estimation of the
single-group model, 
leading to a clear advantage in these conditions.

@@

In this simulation, we demonstrate that the proposed statistics are 
relatively insensitive to minor parameter violations that are
unrelated to the ordering of the auxiliary variable.  As noted
earlier, this feature is especially applicable to situations where
one's sample size is very large.  Analysts often resort to informal
fit measures in practice, because the traditional LRT is nearly
guaranteed to result in significance.  This simulation is intended to
show that the proposed ordinal tests remain viable for large $n$.

@@

\citeA{FroFan11} used confirmatory factor 
models to study the invariance of three youth gratitude scales across
students aged 10 to 19 years.  Due to sample size constraints, the age
variable included six categories: 10--11 years, 12--13 years, 14 years,
15 years, 16 years, and 17--19 years.  Thus, age is an ordinal variable
to which the proposed tests can be applied.
